Comparing \pkg{binsegRcpp} with other implementations of binary segmentation for change-point detection

### 2025-05-08 article.Rnw
done paragraphs based on table 1

%% - \Abstract{} almost as usual
\Abstract{
  Binary segmentation is a classic algorithm for detecting
  change-points in sequential data. In theory, using a simple loss
  function like Normal or Poisson, binary segmentation should be
  extremely fast for $N$ data and $K$ segments: asymptotically $O(N K)$ time in the worst
  case, and $O(N \log K)$ time in the best case.
  In practice,
  existing implementations can be asymptotically slower, and can 
  return incorrect results.
  We propose \pkg{binsegRcpp}, an \proglang{R} package which
  provides a correct \proglang{C++} implementation,
  with the expected asymptotic time complexity.
  %% In practice, existing
  %% implementations such as \pkg{ruptures} (\proglang{Python} module) and
  %% \pkg{changepoint} (\proglang{R} package) are asymptotically slower.
  %% Additionally, \pkg{changepoint} sometimes returns incorrect results.
  %% We present the \proglang{R} package \pkg{binsegRcpp}, which provides an efficient
  %% and correct C++ implementation of binary segmentation.
  We discuss several important \proglang{C++} coding techniques,
  and include detailed comparisons with other implementations of binary
  segmentation:
  \pkg{ruptures}, \pkg{fpop}, \pkg{changepoint}, \pkg{blockcpd}, and
  \pkg{wbs}.
}

@@

Change-point detection is an important problem in sequential data
analysis, that occurs in a variety of different applications, such as
genomics \citep{Hocking2013bioinformatics}, neuroscience
\citep{Jewell2019biostatistics}, and medical monitoring
\citep{Fotoohinasab2020segmentationEMBC}, among others.
The focus of
this article is change-point detection in the off-line setting, in
which we assume there is a sequence of $N$ observed data, and we want
an algorithm to identify the precise locations of any abrupt changes
that may be present \citep{Truong2020}. We assume there is
user-specified maximum number of change-points $K$, and we would like
to compute a sequence of models from 1 to $K$ change-points.

Using
the classical dynamic programming algorithm of
\citet{segment-neighborhood}, optimal change-points for a given cost
function, and a sequence of model sizes $k\in\{1,\dots,K\}$, can be
computed in $O(N^2 K)$ time.
%\citep{Jackson2005}
More recently, the pruned dynamic programming algorithms of
\citet{Maidstone2017} can be used to compute the same sequence of $K$
optimal models, with the same $O(N^2 K)$ worst case time complexity,
but faster $O(KN)$ time best case time complexity.  Other dynamic
programming algorithms such as PELT \citep{pelt} and DuST
\citep{truong2024efficient} compute a single model in best case linear
time $O(N)$, worst case quadratic time $O(N^2)$.  They may be used as
a sub-routine to efficiently compute a range of $K$ change-point
models, using the CROCS algorithm \citep{Liehrmann2021chipseq}, which
again results in best case $O(N K)$ time, worst case $O(N^2 K)$ time.

Binary segmentation is a heuristic algorithm for computing a sequence of models from 1 to $K$ change-points \citep{binary-segmentation}.
Using a simple loss function like the square loss, the best case of binary segmentation is $O(N \log K)$ time \citep{Hocking2024finite}, which is much faster than dynamic programming.
The worst case of binary segmentation is $O(N K)$ time, which is the same speed as the best case of dynamic programming.

\begin{table}[t!]
  \centering
  \small
\begin{tabular}{l|cccccc}
%   > ruptures$version$version
% [1] "1.1.6"
% > sapply(c("binsegRcpp", "changepoint", "wbs", "fpop"), function(p)paste(packageVersion(p)))
%  binsegRcpp changepoint         wbs        fpop 
% "2022.3.29"     "2.2.3"       "1.4" "2019.8.26"
% https://github.com/HaotianXu/changepoints does not have binary segmentation.
  package & binsegRcpp & changepoint & wbs & fpop & ruptures & blockcpd\\
  function & binseg & cpt.mean & sbs & multiBinSeg & Binseg & fit\_blockcpd\\
  version & 2022.3.29 & 2.2.3 & 1.4 & 2019.8.26 & 1.1.9 & 1.0.0\\
\hline
weights & yes & no & no & no & no & no\\
min len & yes & yes & no & no & yes & yes\\
max segs & yes & yes & no & yes & yes & yes\\
dim & one & one & one & multi & multi & multi \\
correct & yes & no & yes & yes & yes & yes\\
losses & 5 & 6 & 1 & 1 & 10 & 5\\
language & C++ & C & C & C++ & Python & C++\\
% all propose some loss computation.
storage & heap & arrays & recursion & heap & LRU cache & heap \\
space & $O(S)$ & $O(S^2)$ & $O(S)$ & $O(S)$ & $O(S)$ & $O(S)$ \\
cumsum & yes & yes & yes & yes & no & yes\\
best & $O(N\log N)$ & $O(N^3)$ & $O(N\log N)$ & $O(N\log N)$ & $O(N^{1.4})$ & $O(N\log N)$\\
worst & $O(N^2)$ & $O(N^3)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$\\
CV & yes & no & no & no & no & no\\
params & all & one & no & no & no &one\\
\end{tabular}
\caption{\label{tab:packages}
  Properties (rows) of six binary segmentation libraries (columns).
Properties are weights (can specify observation-specific positive weights for the loss), min len (can specify minimum segment length parameter), max segs (can specify maximum number of segments parameter), dim (dimensions of input data), correct (does code return correct model parameters?), losses (number of loss functions implemented), language (programming language to implement loss), storage (technique used to store cost of splitting a segment), space (space complexity in number of segments $S$), cumsum (cumulative sum used), best (time complexity in best case data), worst (time complexity in worst case data), CV (support for cross-validation), params (parameters returned, one or all model sizes). }
\end{table}

\subsection{Existing implementations of binary segmentation}
There are several free/open-source implementations of binary
segmentation, most of which use C/C++ code for efficiency
(Table~\ref{tab:packages}).
We compared our proposed implementation to these baselines in terms of functionality, correctness, and efficiency.

The \pkg{fpop} package in R contains the \code{fpop::multiBinSeg}
function, which implements the square loss for multi-dimensional data
\citep{Maidstone2017}.

The \pkg{wbs} package in R contains the \code{wbs::sbs} function,
which implements the square loss for univariate data
\citep{Baranowski2019}.  Since it uses recursive function calls, it
does not support the max segments parameter $K$, and instead always
computes the full path of models (from 1 to $N-1$ change-points).

The \code{changepoint::cpt.mean(method="BinSeg")} \citep{Killick2014,Killick2022}.

TODO \code{ruptures.Binseg} \citep{Truong2020}.

TODO \code{fastcpd} package \citep{fastcpd2024} (nice related work table).

TODO \pkg{blockcpd} package TODO \citep{blockcpd2021,Prates2021} .

\section{Models and software} \label{sec:models}

@@

\paragraph{Recursive update rules.}
Let $K\in\{ 2, \dots, N\}$ be the user-specified max number of segments to consider.
The binary segmentation algorithm recursively computes the following quantities for each $k\in\{2,\dots,K\}$.
First, the best segment to split is computed by greedy minimization of the loss decrease over all splittable segments.
\begin{equation}
  j^*_k, e^*_k, d^*_k, t^*_k = \argmin_{(j,e,d,t)\in\mathcal S_{k-1}} d.
\end{equation}
Ideally, the minimization above takes $O(1)$ constant time, if the set of splittable segments is a red-black tree (C++ STL multiset).
As a baseline, we also consider a linked list, which results in a sub-optimal $O(k)$ linear time search (see Section~\ref{sec:data-structures} for details).

@@

In general, for any number of segments $K$, we can compute a table of
estimated means (one row for each segment), using the following
algorithm.  First, consider only the first $K$ rows of the
\code{splits} table.  Next, set values in \code{before.mean} and
\code{after.mean} columns to missing (\code{NA}), based on values in
\code{invalidates.index} and \code{invalidates.after} columns.
Finally, sort the \code{end} values, and use the corresponding
non-missing mean values.
For $K$ segments, this algorithm is $O(K \log K)$ time due to the sort.
It is implemented in the \code{coef} method, as shown in the example code below:

>>>>

%% - \Abstract{} almost as usual
\Abstract{ Binary segmentation is a classic algorithm for detecting
  change-points in sequential data. In theory, using a simple loss
  function like Normal or Poisson, binary segmentation should be
  extremely fast for $N$ data and $S$ segments: asymptotically $O(N
  S)$ time in the worst case, and $O(N \log S)$ time in the best case.
  In practice, existing implementations can be asymptotically slower,
  and can return incorrect results.  We propose \pkg{binsegRcpp}, an
  \proglang{R} package which provides a correct \proglang{C++}
  implementation, with the expected asymptotic time complexity.
  %% In practice, existing
  %% implementations such as \pkg{ruptures} (\proglang{Python} module) and
  %% \pkg{changepoint} (\proglang{R} package) are asymptotically slower.
  %% Additionally, \pkg{changepoint} sometimes returns incorrect results.
  %% We present the \proglang{R} package \pkg{binsegRcpp}, which provides an efficient
  %% and correct C++ implementation of binary segmentation.
  We discuss several important \proglang{C++} coding techniques,
  and include detailed comparisons with other implementations of binary
  segmentation:
  \pkg{ruptures}, \pkg{fpop}, \pkg{changepoint}, \pkg{blockcpd}, and
  \pkg{wbs}.
}

@@

Change-point detection is an important problem in sequential data
analysis, that occurs in a variety of different applications, such as
genomics \citep{Hocking2013bioinformatics}, neuroscience
\citep{Jewell2019biostatistics}, and medical monitoring
\citep{Fotoohinasab2020segmentationEMBC}, among others.
The focus of
this article is change-point detection in the off-line setting, in
which we assume there is a sequence of $N$ observed data, and we want
an algorithm to identify the precise locations of any abrupt changes
that may be present \citep{Truong2020}. We assume there is
user-specified maximum number of segments $S$, and we would like
to compute a sequence of models from 1 to $S$ segments (0 to $S-1$ change-points).

Using
the classical dynamic programming algorithm of
\citet{segment-neighborhood}, optimal change-points for a given cost
function, and a sequence of model sizes $k\in\{1,\dots,S\}$, can be
computed in $O(N^2 S)$ time.
%\citep{Jackson2005}
More recently, the pruned dynamic programming algorithms of
\citet{Maidstone2017} can be used to compute the same sequence of $S$
optimal models, with the same $O(N^2 S)$ worst case time complexity,
but faster $O(NS)$ time best case time complexity.  Other dynamic
programming algorithms such as PELT \citep{pelt} and DuST
\citep{truong2024efficient} compute a single model in best case linear
time $O(N)$, worst case quadratic time $O(N^2)$.  They may be used as
a sub-routine to efficiently compute a range of $S$ change-point
models, using the CROCS algorithm \citep{Liehrmann2021chipseq}, which
again results in best case $O(N S)$ time, worst case $O(N^2 S)$ time.
\citet{fastcpd2024} propose a change-point algorithm based on dynamic programming, and provide comparisons with several previous algorithms.

Binary segmentation is a heuristic algorithm for computing a sequence of models from 1 to $S$ change-points \citep{binary-segmentation}.
Using a simple loss function like the square loss, the best case of binary segmentation is $O(N \log S)$ time \citep{Hocking2024finite}, which is much faster than dynamic programming.
The worst case of binary segmentation is $O(N S)$ time, which is the same speed as the best case of dynamic programming.

\begin{table}[t!]
  \centering
  \small
\begin{tabular}{l|cccccc}
%   > ruptures$version$version
% [1] "1.1.6"
% > sapply(c("binsegRcpp", "changepoint", "wbs", "fpop"), function(p)paste(packageVersion(p)))
%  binsegRcpp changepoint         wbs        fpop 
% "2022.3.29"     "2.2.3"       "1.4" "2019.8.26"
% https://github.com/HaotianXu/changepoints does not have binary segmentation.
  package & binsegRcpp & changepoint & wbs & fpop & ruptures & blockcpd\\
  function & binseg & cpt.mean & sbs & multiBinSeg & Binseg & fit\_blockcpd\\
  version & 2022.3.29 & 2.2.3 & 1.4 & 2019.8.26 & 1.1.9 & 1.0.0\\
\hline
weights & yes & no & no & no & no & no\\
min len & yes & yes & no & no & yes & yes\\
max segs & yes & yes & no & yes & yes & yes\\
dim & one & one & one & multi & multi & multi \\
correct & yes & no & yes & yes & yes & yes\\
losses & 5 & 6 & 1 & 1 & 10 & 5\\
language & C++ & C & C & C++ & Python & C++\\
% all propose some loss computation.
storage & heap & arrays & recursion & heap & LRU cache & heap \\
space & $O(S)$ & $O(S^2)$ & $O(S)$ & $O(S)$ & $O(S)$ & $O(S)$ \\
cumsum & yes & yes & yes & yes & no & yes\\
best & $O(N\log N)$ & $O(N^3)$ & $O(N\log N)$ & $O(N\log N)$ & $O(N^{1.4})$ & $O(N\log N)$\\
worst & $O(N^2)$ & $O(N^3)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$\\
CV & yes & no & no & no & no & no\\
params & all & one & no & no & no &one\\
\end{tabular}
\caption{\label{tab:packages} Properties (rows) of six binary
  segmentation libraries (columns).  Properties are weights (can
  specify observation-specific positive weights for the loss), min len
  (can specify minimum segment length parameter), max segs (can
  specify maximum number of segments parameter), dim (dimensions of
  input data), correct (does code return correct model parameters?),
  losses (number of loss functions implemented), language (programming
  language to implement loss), storage (technique used to store cost
  of splitting a segment), space (space complexity in number of
  segments $S$), cumsum (cumulative sum used), best and worst case
  time complexity (using $N$ data and $S=N$ segments), CV (support for
  cross-validation), params (parameters returned, one or all model
  sizes). }
\end{table}

\subsection{Existing implementations of binary segmentation}
There are several free/open-source implementations of binary
segmentation, most of which are \proglang{R} packages with
\proglang{C}/\proglang{C++} code (Table~\ref{tab:packages}).  We
compared our proposed implementation to these baselines in terms of
functionality, correctness, and efficiency.  None of these existing
implementations support cross-validation, nor observation-specific
weights, which are unique features that we propose in \pkg{binsegRcpp}.

The \pkg{changepoint} package in \proglang{R} contains the
\code{cpt.mean} function which implements several
different change-point detection algorithms, including binary
segmentation via \code{method="BinSeg"}
\citep{Killick2014,Killick2022}.  Six loss functions are supported,
along with min segment length parameter, max segments parameter, and
returning the segment-specific parameters for one model (not all).
The
implementation does not always return correct results, and requires
more time/space than expected. For $S$ segments, an $S\times S$ matrix
is used to represent the segment end positions, whereas only a vector
of $S$ values is required. Therefore, the time complexity is $O(N
S^2)$, much slower than the expected $O(N\log S)$. Likewise, the space
complexity is $O(S^2)$ rather than the expected $O(S)$.

The \pkg{wbs} package in \proglang{R} contains the \code{sbs} function,
which implements the square loss for univariate data
\citep{Baranowski2019}. Since it uses recursive function calls, it
does not support the max segments parameter $S$, and instead always
computes the full path of models (from 1 to $N-1$ change-points).
It does not implement min segment length, nor
returning the segment mean parameters.

The \pkg{fpop} package in \proglang{R} contains the \code{multiBinSeg}
function, which implements the square loss for multi-dimensional data
\citep{Maidstone2017}. It implements the max segments parameter, but
not min segment length, nor
returning the segment mean parameters.

The \pkg{ruptures} package in \proglang{python} contains the
\code{Binseg} class, which implements binary segmentation for 10 loss
functions \citep{Truong2020}. It supports min length parameter and max
segments parameter. The cumulative sum is not used in the
implementation of the square loss, which is normally required to
achieve optimal best case time complexity. When running on $N$ best
case data, with $S=N$ segments, we have observed approximately
$O(N^{1.4})$ time complexity (see Results section), which is slower
than the expected/optimal $O(N\log N)$ time complexity.

The \pkg{blockcpd} package in \proglang{R} contains the
\code{fit\_blockcpd} function, which implements several algorithms for
change-point detection, including binary segmentation via
\code{method="hierseg"} \citep{blockcpd2021,Prates2021}. It supports
input of min segment length and maximum segments, and output of
segment mean parameters for one model size (not all).

\section{Statistical model and algorithm} \label{sec:models}

@@

\paragraph{Recursive update rules.}
Let $S\in\{ 2, \dots, N\}$ be the user-specified max number of segments to consider.
The binary segmentation algorithm recursively computes the following quantities for each $k\in\{2,\dots,S\}$.
First, the best segment to split is computed by greedy minimization of the loss decrease over all splittable segments.
\begin{equation}
  j^*_k, e^*_k, d^*_k, t^*_k = \argmin_{(j,e,d,t)\in\mathcal S_{k-1}} d.
\end{equation}
Ideally, the minimization above takes $O(1)$ constant time, if the set of splittable segments is a red-black tree (C++ STL multiset).
As a baseline, we also consider a linked list, which results in a sub-optimal $O(k)$ linear time search (see Section~\ref{sec:data-structures} for details).

@@

In general, after running binary segmentation up to $S$ segments, we
can compute a table of estimated means (one row for each segment), for
any number of segments $k\in\{1,\dots,S\}$, using the following algorithm.
First, consider only the first $k$ rows of the \code{splits} table.
Next, set values in \code{before.mean} and \code{after.mean} columns
to missing (\code{NA}), based on values in \code{invalidates.index}
and \code{invalidates.after} columns.  Finally, sort the \code{end}
values, and use the corresponding non-missing mean values.  For $k$
segments, this algorithm is $O(k \log k)$ time due to the sort.  It is
implemented in the \code{coef} method, as shown in the example code
below:
